Select-based random access to variable-byte encodings

### 2020-02-28 gradu/msc_english_contents.tex
Moved algorithm
centered 2.1
Fixes

VB encoding splits each integer into blocks of $b$ bits and adds a continuation bit to the block to form chunks of length $1+b$. The extra bit is set to 1 only
on the block containing the least significant bits of the integer. This information is used in decoding to signal if next chunk continues the current 
integer. For example, let's assume $b = 4$ and let $n = 42$ be a 16 bit unsigned integer. The standard 8-bit representation of $n$ is 
\texttt{00000000 00101000}. When split to blocks of $b$ bits, it becomes \texttt{0000 0000 0010 1000}. Empty blocks are omitted and continuation bits 
are added to the remaining blocks. The result is \texttt{00010 11000}, which is the compressed data.
\begin{figure}[ht]
\begin{algorithmic}
\Function{VBEncodeNumber}{$n$}
\State $bytes\gets $ list
\While{true}
\State \Call{prepend}{bytes,$n \bmod 128$}
\If{$n < 128$} \State break \EndIf
\State $n\gets n$ div $128$
\EndWhile
\State $bytes$[\Call{len}{$bytes$}-1] += $128$
\State \textbf{return} bytes
\EndFunction

\Function{VBEncode}{$numbers$}
\State $bytestream\gets $ list
\ForEach {$n \in numbers$}
\State $bytes \gets$ VBEncodeNumber($n$)
\State $bytestream$.extend($bytes$)
\EndFor
\State \textbf{return} bytestream
\EndFunction

\end{algorithmic}
\captionof{figure}{VByte encoding} \label{vbyte_enc}
\end{figure}

Decoding is essentially just reversing the encoding steps: chunks are read until a chunk with 1 as continuation bit is found. Continuation bits 
are removed from all the chunks and the blocks are concatenated to form the original number. As in the previous example, $b = 4$, encoded message 
is \texttt{00010 11000} and $n = 0$. Block from the first chunk is extracted and added to $n$, making $n = $\texttt{10}. A bitwise shift to left 
equal to $b$ is applied to $n$, changing $n = $\texttt{100000}. Then the block is extracted from the next chunk. This block is added  $n$, making 
$n = 42$. Because the previous continuous bit was 1, decoding this number has finished. An example implementation of encode and decode with block 
length of 7 is shown in Figure~\ref{vbyte_enc} and Figure~\ref{vbyte_dec}.

\begin{figure}[ht]
\begin{algorithmic}
\Function{VBDecode}{$bytestream$}
\State $numbers\gets $ list
\State $n\gets 0$
\ForEach {$b \in bytestream$}
\If{$b < 128$}
\State $n\gets 128\times n $ + $b$
\Else
\State $n\gets 128\times n $ + $b$ - $128$
\State $numbers$.append($n$)
\State $n\gets 0$
\EndIf
\EndFor
\State \textbf{return} numbers
\EndFunction
\end{algorithmic}
\caption{VByte decoding} \label{vbyte_dec}
\end{figure}

Small lengths of $c$ can yield better compression rate at the cost of more bit manipulation, while longer chunks need less bit manipulation and 
offer less effective compression. Generally block length of 7 has been used because it gives a good average and handling chunks as bytes is 
convenient \citep{Man08}.

VB encoding is a well known compression algorithm. It's origins date back to 1980's and the famous MIDI music file format. It stored some of the numbers
in a "variable-length quantity" form, which was a 7-bit block VB structure \citep{Mid96}. Similar data types have existed for example in Apache Lucene 
(as vInt) and IBM DB2 database (as Variable Byte). Later, VB encoding was found efficient in compressing integer lists. It was first experimented in 
compressing inverted index lists of word locations in documents \citep{Sch02}. That yielded excellent records, and since then many different approaches 
have been introduced.

Modern studies have looked into machine code for VB and applied SIMD (Single instruction, multiple data) architecture to VB \citep{Lem18,Pla15}. The bit 
operations in VB are simple and therefore modifying the code to use SIMD instructions is straightforward and the speed improvements are significant.

TODO: might need 1-2 more chapters

@@

Ability to handle large amounts of data fast is one of the key challenges in the field of search engines. Compressed data structures are applied to fit the 
data into cache, memory or even hard drive. Direct access to any element is one of the usual requirements in compressed data strutures. It is not natively
possible to decode the i-th element in variable-byte compression algorithms, because the position of the element depends on the length of the preceding 
compressed data.

Direct access is achievable with supporting data structures. An obvious solution is to store the location of each element, but it adds a very large overhead 
which removes the benefit of compression.

\section{Rank and Select}
Rank and select are two succinct data structure operations which operate on a bit array B. $Rank_1$(B, i) gives the sum of 1 bits between B[0] and B[i] and 
$select_1$(b, i) gives the index of i-th 1 bit in B. Both operations work in constant time (citation?). Locations of 1 bits in B should represent locations 
of the items in the encoded data. For most compression algorithms, this requires B to be created in addition to the existing data and the length of B usually 
has to be close to the length of the data. VB encoding has several advantages with search and rank: it's data is compressed in blocks of equal length which 
significantly shrinks the length of B. Also the bit array C formed from the continuation bits already stores the locations of items. In this case, $rank_1$(C, i) 
would give the sum of end bytes up to i-th index and $select_1$(C, i) would give the location of the ending byte of i-th compressed element.

@@

\section{DAC via rank}
Directly addressable codes in VB was first introduced by \citep{Bri09} in 2009. Their solution was to divide chunks in encoded bytes to separate arrays $A_1$ to 
$A_n$ and then use $rank_0$ to get direct access. If i is the requested index, then $rank_0$($A_k$, i) returns the number of elements preceding i that have 
continuation bit set at 0. In other words, the number of elements that continue to $A_{k+1}$.

@@

To use $select_1$ with VB, continuation bits need to be separated from chunks to their own bit array and a $select_1$ structure built over it. $Select$(i) returns the
location of the end byte of i-th element. Therefore the start of j-th element in the block array is at block $b_s$ = $select_1$(j-1) + 1. To simplify calculations even further,
block sizes of 4 and 8 were used to avoid splitting of blocks between bytes.

Unlike the standard VB encoding, the continuation bits are not in the middle of the data blocks. The compressed number can be read from the memory block and concatenation of blocks
is not needed. The data in blocks is be written into memory as it is, so that when reading a large unsigned integer from the block byte array, the bits are already in correct order.

If the original element size is 32 bits and the data was compressed to k blocks of length b. The starting byte s of the element is $select$(i) * b div 8 where div is the integer division. 
Offset o is the remainder of previous division, o = x*b mod 8. The 32-bit value is read from memory from byte location s. Then bit shift left for 32-b*k-o is applied to remove trailing 
bits and bit shift right 32-b*k to re-align bits. If block size is one byte, offset calculation is not needed and thus the process is slightly faster.

The intuitive way to calculate block length k of i-th element is from $select_1$(i+1) and subtract previously calculated start block index from it. This however causes a second 
$select_1$ query, which is costly. A much faster option is to iterate forward in bit array until the next 1. Alternatively, the block length can be calculated by reading an integer 
from the bit array, aligning it's offset and counting the trailing zeros. Figure~\ref{select_pseudo} contains an example of VB encoding with DAC with $select$ and block size 8. 
Different block sizes need extra calculation to get the block location from the byte array. In the example, $CalculateLength$ returns the length of the number in blocks and $ByteMask(k)$ returns 
a bit mask for k bytes.

(maybe to future work) Storing bytes in this fashion may have benefits when reading subsequent (fast next() and previous()?) elements. With some modifications, SIMD instructions can possibly be applied.

\begin{figure}[ht]
\begin{algorithmic}
\State $i \gets $ wanted index
\State $A \gets $ block byte array
\State $B \gets $ continuation bit array
\State $begin \gets \Call{Select}{index}$
\State $len \gets \Call{CalculateLength}{begin}$
\State $mask \gets \Call{ByteMask}{len}$
\State $number \gets A[begin]$ \Comment{read a larger number from the byte array}
\State $number \gets number \mathbin{\&} mask$

@@

TODO: try bitvector SD performance + size (select)

TODO: look into select implementation, may need optimizing

>>>>

VB encoding splits each integer into blocks of $b$ bits and adds a continuation bit to the block to form chunks of length $1+b$. The extra bit is set to 1 only
on the block containing the least significant bits of the integer. This information is used in decoding to signal if next chunk continues the current 
integer. For example, let's assume $b = 4$ and let $n = 42$ be a 16 bit unsigned integer. The standard 8-bit representation of $n$ is 
\texttt{00000000 00101000}. When split to blocks of $b$ bits, it becomes \texttt{0000 0000 0010 1000}. Empty blocks are omitted and continuation bits 
are added to the remaining blocks. The result is \texttt{00010 11000}, which is the compressed data.

\begin{figure}[ht]
\centering
  \begin{minipage}{0.5\linewidth}
\begin{algorithmic}[H]
\Function{VBEncodeNumber}{$n$}
\State $bytes\gets $ list
\While{true}
\State \Call{prepend}{bytes,$n \bmod 128$}
\If{$n < 128$} \State break \EndIf
\State $n\gets n$ div $128$
\EndWhile
\State $bytes$[\Call{len}{$bytes$}-1] += $128$
\State \textbf{return} bytes
\EndFunction
\medskip
\medskip
\Function{VBEncode}{$numbers$}
\State $bytestream\gets $ list
\ForEach {$n \in numbers$}
\State $bytes \gets$ \Call{VBEncodeNumber}{$n$}
\State \Call{extend}{$bytestream$,$bytes$}
\EndFor
\State \textbf{return} bytestream
\EndFunction

\end{algorithmic}
\end{minipage}
\captionof{figure}{VByte encoding} \label{vbyte_enc}
\end{figure}

Decoding is essentially just reversing the encoding steps: chunks are read until a chunk with 1 as continuation bit is found. Continuation bits 
are removed from all the chunks and the blocks are concatenated to form the original number. As in the previous example, $b = 4$, encoded message 
is \texttt{00010 11000} and $n = 0$. The block from the first chunk is extracted and added to $n$, making $n = $\texttt{10}. A bitwise shift to the left 
equal to $b$ is applied to $n$, changing $n = $\texttt{100000}. Then the block is extracted from the next chunk. This block is added  $n$, making 
$n = 42$. Because the previous continuation bit was 1, decoding for this number is complete. More examples shown in Table~\ref{table:vbytes}. 
An example implementation of encode and decode with block length of 7 is shown in Figure~\ref{vbyte_enc} and Figure~\ref{vbyte_dec}.

\begin{figure}[ht]
\centering
  \begin{minipage}{0.5\linewidth}
\begin{algorithmic}[H]
\Function{VBDecode}{$bytestream$}
\State $numbers\gets $ list
\State $n\gets 0$
\ForEach {$b \in bytestream$}
\If{$b < 128$}
\State $n\gets 128\times n $ + $b$
\Else
\State $n\gets 128\times n $ + $b$ - $128$
\State $numbers$.append($n$)
\State $n\gets 0$
\EndIf
\EndFor
\State \textbf{return} numbers
\EndFunction
\end{algorithmic}
\end{minipage}
\caption{VByte decoding} \label{vbyte_dec}
\end{figure}

Small lengths of $c$ can yield better compression rate at the cost of more bit manipulation, while longer chunks need less bit manipulation and 
offer less effective compression. Generally block length of 7 has been used because it tends to perform well on average and handling chunks as bytes is 
convenient \citep{Man08}.

VB encoding is a well known compression algorithm. It's origins date back to 1980's and the famous MIDI music file format. It stored some of the numbers
in a "variable-length quantity" form, which was a 7-bit block VB structure \citep{Mid96}. Similar encoding types have existed for example in Apache Lucene 
(as vInt) and IBM DB2 database (as Variable Byte). Later, VB encoding was was to be found efficient in compressing integer lists. It was first experimented 
as a tool for compressing inverted index lists of word locations in documents \citep{Sch02}. That yielded excellent records, and since then many different 
approaches have been introduced.

More recent studies have looked into machine code for VB and applied SIMD (Single instruction, multiple data) instructions to VB \citep{Lem18,Pla15}. The bit 
operations in VB are simple and therefore modifying the code to use SIMD instructions is straightforward and the speed improvements are significant.

\begin{table}
\centering
\begin{tabular}{l||c c c c c} 
Original number & first block & second block & third block & fourth block &\\ 
\hline \hline 
4  & \texttt{\underline{1}4000}    &                             &                           &  &  \\
17  & \texttt{\underline{0}0001}   & \texttt{\underline{1}0001}  &                           &  &  \\
620  & \texttt{\underline{0}1100}  & \texttt{\underline{0}0110} & \texttt{\underline{1}0010} &  &  \\
60201 & \texttt{\underline{1}1001} & \texttt{\underline{0}0010} & \texttt{\underline{0}1011} & \texttt{\underline{1}1110} &  \\

\hline
\end{tabular}
\caption{VByte encoded numbers, block size 4. Continuation bit underlined.\label{table:vbytes}}
\end{table}

@@

The ability to handle large amounts of data fast is one of the key challenges in the field of search engines. Compressed data structures are applied to fit the 
data into cache, memory or even hard drive. Direct access to any element in a compressed list or array is one of the usual requirements in compressed data 
strutures. It is not natively possible to decode the i-th element in variable-byte compression algorithms, because the position of the element in the compressed
list depends on the length of the preceding compressed data. Direct access is achievable with supporting data structures. A naive solution is to store the location 
of each element, but it adds a very large overhead which removes the benefit of compression.

\section{Rank and Select}
Rank and select are two succinct data structure operations which operate on a bit array B. $Rank_1$(B, i) gives the sum of 1 bits between B[0] and B[i] and 
$select_1$(b, i) gives the index of i-th 1 bit in B. Both operations can be implemented to work in constant time \citep{gbmp2014sea}. Locations of 1 bits in B should represent locations 
of the items in the encoded data. For most compression algorithms, this requires B to be created in addition to the existing data and the length of B usually 
has to be close to the length of the data. VB encoding has several advantages with search and rank: it's data is compressed in blocks of equal length which 
significantly shrinks the length of B. Also the bit array C formed from the continuation bits already stores the locations of items. In this case, $rank_1$(C, i) 
would give the sum of end bytes up to i-th index and $select_1$(C, i) would give the location of the ending byte of i-th compressed element.

@@

\section{DAC via rank}
Directly addressable codes in VB was first introduced by \citep{Bri09} in 2009. Their solution was to divide chunks in encoded bytes to separate arrays $A_1$ to 
$A_n$ and then use $rank_0$ to get direct access. If i is the requested index, then $rank_0$($A_k$, i) returns the number of elements preceding i that have 
continuation bit set at 0. In other words, the number of elements that continue to $A_{k+1}$. See Figure~\ref{bris_ds} for visualization of data.

\begin{figure}[H]
\centering
\includegraphics[scale=0.4]{bris.png}
\caption{Data structure by Brisaboa et al., visualized} \label{bris_ds}
\end{figure}

@@

To use $select_1$ with VB, continuation bits need to be separated from chunks to their own bit array and a $select_1$ structure built over it. Because every compressed element has 1
only on it's last block, $Select_1$(i) returns the location of the end byte of i-th element. Therefore the start of j-th element in the block array is at 
block $b_s$ = $select_1$(j-1) + 1. Implementation was simplified by using only block sizes 4 and 8 to prevent block splitting between bytes.

Unlike the standard VB encoding, the continuation bits are removed from the chunks and stored in their own array, which leaves the blocks in their own array. This allows the 
compressed number to be read from the memory block and removes the need of block concatenation. The data in blocks is written into memory as they appear in the original integer, so that 
when reading a word from the block byte array, the bits and bytes are already in correct order.

If the original element size is 32 bits and the data was compressed to k blocks of length b. The starting byte s of the element is $select$(i) * b div 8 where div is the integer division. 
Offset o is the remainder of previous division, o = x*b mod 8. The 32-bit word is read from memory from byte location s. Then bit shift left for 32-b*k-o is applied to remove trailing 
bits and bit shift right 32-b*k to re-align bits. If block size is one byte, offset calculation is not needed and thus the process is slightly faster.

The most intuitive way to calculate block length of i-th element is from $select_1$(i+1) and subtract the previously calculated start block index. This however causes a second 
$select_1$ query, which is costly. A much faster option is to iterate forward in bit array until the next 1. Alternatively, the block length can be calculated by reading an integer 
from the bit array, aligning it's offset and counting the trailing zeros. Figure~\ref{select_pseudo} contains an example of VB decoding with DAC with $select$ and block size 8. 
Different block sizes need extra calculation to get the block location from the byte array. In the example, $CalculateLength$ returns the length of the number in blocks and $ByteMask(k)$ returns 
a bit mask for k bytes.

\begin{figure}[ht]
\centering
\begin{algorithmic}
\State $i \gets $ wanted index
\State $A \gets $ block byte array
\State $B \gets $ continuation bit array
\State $begin \gets \Call{Select}{index}$
\State $len \gets \Call{CalculateLength}{begin}$
\State $mask \gets \Call{ByteMask}{len}$
\State $word \gets A[begin]$ \Comment{read a word from the byte array}
\State $word \gets word \mathbin{\&} mask$

@@

Figure~\ref{VBarrays} portrays an example of the bit array and the block array. $Select_1(k-1)$+1 returns the index of the starting byte of ith compressed element. Length of the compressed 
element is obtainable e.g. via $select_1(k)$. Then a word is read from the block array, starting from block $A_i$. Because element length in this case is 3, blocks not belonging to the element
are removed either with a bitmask or by double bit shifting, resulting in decoded k-th element.

\begin{figure}[ht]
\centering
\begin{tikzpicture}
[node distance=0pt,
 start chain = A going right,
 start chain = B going right,
 arrow/.style = {draw=#1,-{Stealth[]}, 
                shorten >=1mm, shorten <=1mm}, % styles of arrows
 arrow/.default = black,
 X/.style = {rectangle, draw,% styles of nodes in string (chain)
                minimum width=9ex, minimum height=5ex,
                outer sep=0pt, on chain}
 ]

\node[X, join] (start) {...};
\node[X] (first){0};
\node[X, very thick] (select){1};
\node[X] (Btarget){0};
\node[X, join] (second){0};
\node[X, very thick, join] (end){1};
\node[X, join] (third){0};
\node[X, join] {$B_{...}$};

\node [above=1ex of start]{Bit array};
\node [below=1ex of start]{index:};
\node [above=1ex of select]{k-1th 1};
\node [below=1ex of first]{$i-1$};
\node [below=1ex of select]{$i$};
\node [below=1ex of Btarget]{$i+1$};
\node [below=1ex of second]{$i+2$};
\node [below=1ex of end]{$i+3$};
\node [below=1ex of third]{$i+4$};
\node [above=1ex of end]{kth 1};
\node[X, below=12ex of start] (x) {...};
\node[X, join] {$A_{i-2}$};
\node[X, join] {$A_{i-1}$};
\node[X, ultra thick] (Atarget){$A_{i}$};
\node[X, ultra thick] {$A_{i+1}$};
\node[X, ultra thick] {$A_{i+2}$};
\node[X, join] {$A_{i+3}$};
\node[X, join] {$A_{...}$};
\node [above=1ex of x]{Block array};

\draw[arrow] (Btarget.south) to (Atarget.north);
\end{tikzpicture}
\caption{Data} \label{VBarrays}
\end{figure}
TODO: try bitvector SD performance + size (select)
